---
title: "The idionomics Pipeline: Individual-First Time-Series Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The idionomics Pipeline: Individual-First Time-Series Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

## The idionomic principle

Classical panel methods (multilevel models, fixed-effects regression) estimate one set of parameters shared across all subjects. This is only sensible under strong assumptions such as ergodicity and exchangeability. When those assumptions fail — and they often do in psychological and behavioral data — group averages may obscure substantial individual differences: a positive average effect can coexist with a meaningful subset of individuals for whom the effect is negative, zero, or nonsignificant.

**idionomics** inverts the order of operations:

1. **Individual first.** Fit a model to each person's time series independently, capturing that person's unique dynamics and effect sizes.
2. **Group later.** Aggregate individual estimates using random-effects meta-analysis, which explicitly represents and quantifies between-person heterogeneity.

The recommended pipeline is:

```
i_screener()  →  pmstandardize()  →  i_detrender()  →  iarimax()  →  i_pval() / sden_test()
```

This vignette walks through each step with a synthetic panel of 10 subjects, each contributing 50 time points.

```{r load}
library(idionomics)

set.seed(42)
panel <- do.call(rbind, lapply(1:6, function(id) {
  x <- rnorm(30)
  data.frame(
    id   = as.character(id),
    time = seq_len(30),
    x    = x,
    y    = 0.4 * x + rnorm(30),
    stringsAsFactors = FALSE
  )
}))
```

The true within-person effect of `x` on `y` is 0.4 for all subjects in this example.

---

## Step 1: Data quality screening — `i_screener()`

Before standardizing, screen subjects on their raw data. After `pmstandardize()`, all non-constant series have within-person variance ≈ 1 by construction, making variance-based filters ineffective. `i_screener()` catches low-quality subjects at the right stage.

```{r screener}
panel_clean <- i_screener(
  panel,
  cols    = c("x", "y"),
  idvar   = "id",
  min_sd  = 0.3,
  verbose = TRUE
)
```

Three criteria are available (all optional except `min_n_subject`):

| Criterion | What it catches |
|---|---|
| `min_n_subject` | Too few observations |
| `min_sd` | Near-constant series (floor/ceiling responders) |
| `max_mode_pct` | "Stuck" responders (e.g. ≥ 80% of responses identical) |

Use `mode = "report"` to inspect quality metrics without committing to exclusion:

```{r screener-report}
report <- i_screener(panel, cols = c("x", "y"), idvar = "id",
                     min_sd = 0.3, mode = "report")
print(report)
```

---

## Step 2: Within-person standardization — `pmstandardize()`

Person-mean standardization computes `(x - person_mean) / person_sd` for each subject and column. This removes between-person differences in level and scale, making coefficients comparable across subjects. Output columns are named `<col>_psd`.

```{r pmstandardize}
panel_std <- pmstandardize(panel_clean, cols = c("x", "y"), idvar = "id")
head(panel_std[, c("id", "time", "x_psd", "y_psd")])
```

Edge cases are handled automatically: all-NA series remain NA; single-observation series return 0; constant series (zero SD) return 0 and are filtered downstream by `iarimax()`'s `minvar` guard.

---

## Step 3: Linear detrending — `i_detrender()`

`i_detrender()` removes the within-person linear time trend via `lm(col ~ time)` and replaces each column with the residuals (`<col>_dt`). This reduces the integration order `auto.arima()` selects, keeping models parsimonious.

Filtering is per-column and independent: a subject can produce valid detrended residuals for one variable while receiving `NA` in another if that column fails the variance or observation-count thresholds.

```{r detrender}
panel_dt <- i_detrender(
  panel_std,
  cols    = c("x_psd", "y_psd"),
  idvar   = "id",
  timevar = "time",
  verbose = TRUE
)
head(panel_dt[, c("id", "time", "x_psd_dt", "y_psd_dt")])
```

> **Pipeline order matters.** Detrending after standardization is preferred. Reversing the order (`i_detrender()` → `pmstandardize()`) can amplify near-zero residuals to unit variance, bypassing `iarimax()`'s `minvar` filter.

---

## Step 4: Per-subject ARIMAX and meta-analysis — `iarimax()`

`iarimax()` fits one `auto.arima()` model per subject, extracts coefficients via `broom::tidy()`, and pools the focal predictor's coefficients with `metafor::rma()` (REML).

```{r iarimax}
result <- iarimax(
  panel_dt,
  y_series  = "y_psd_dt",
  x_series  = "x_psd_dt",
  id_var    = "id",
  timevar   = "time"
)
```

### Summary

```{r iarimax-summary}
summary(result)
```

### Caterpillar plot

```{r iarimax-plot, fig.width = 6, fig.height = 5}
plot(result,
     y_series_name = "Outcome (y)",
     x_series_name = "Predictor (x)")
```

Each bar is a per-subject 95% confidence interval, colored green (significantly positive), red (significantly negative), or black (crosses zero). The blue band is the 95% CI of the REMA pooled effect.

### Per-subject results

The full individual-level output is in `$results_df`:

```{r results-df}
head(result$results_df[, c("id", "nAR", "nI", "nMA",
                           "estimate_x_psd_dt", "std.error_x_psd_dt",
                           "n_valid", "n_params", "raw_cor")])
```

---

## Step 5: Per-subject p-values — `i_pval()`

`i_pval()` attaches a `pval_<feature>` column to `results_df` using the two-tailed t-distribution with ML-based degrees of freedom (`n_valid - n_params`). These p-values differ from `lm()` because ARIMA uses `sigma² = SSR/n` rather than `SSR/(n - k)`.

```{r i-pval}
result_pval <- i_pval(result)
result_pval$results_df[, c("id", "estimate_x_psd_dt", "pval_x_psd_dt")]
```

---

## Step 6: Sign Divergence / Equisyncratic Null test — `sden_test()`

`sden_test()` runs a binomial test on the count of individually significant effects. Two variants are available and selected automatically based on the REMA pooled effect:

- **ENT** (Equisyncratic Null Test): tests whether any-direction significant effects exceed the rate expected by chance (`alpha_arimax`). Used when the pooled effect is non-significant — i.e., no dominant direction has been established.
- **SDT** (Sign Divergence Test): tests whether effects in the counter-pooled direction exceed chance (`alpha_arimax / 2`). Used when the pooled effect is significant, as a sensitivity check for hidden heterogeneity.

The auto-selection pivot is fixed at p = 0.05, independent of `alpha_arimax`.

```{r sden}
sden <- sden_test(result_pval)
summary(sden)
```

---

## Full pipeline at a glance

```{r full-pipeline, eval = FALSE}
library(idionomics)

# 1. Screen raw data
panel_clean <- i_screener(panel, cols = c("x", "y"), idvar = "id",
                          min_sd = 0.3)

# 2. Within-person standardize
panel_std <- pmstandardize(panel_clean, cols = c("x", "y"), idvar = "id")

# 3. Detrend
panel_dt <- i_detrender(panel_std, cols = c("x_psd", "y_psd"),
                        idvar = "id", timevar = "time")

# 4. I-ARIMAX
result <- iarimax(panel_dt, y_series = "y_psd_dt", x_series = "x_psd_dt",
                  id_var = "id", timevar = "time")

# 5. Inspect
summary(result)
plot(result, y_series_name = "Outcome", x_series_name = "Predictor")

# 6. Per-subject p-values
result_pval <- i_pval(result)

# 7. SDEN test
sden <- sden_test(result_pval)
summary(sden)
```
